Synthesizing Quotient Lenses

### 2018-03-08 papers/QuotientLenses/main.tex
Added more edits

\newif\ifdraft\drafttrue  % set true to show comments
\newif\ifplentyoftime\plentyoftimefalse  % :-)

@@

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\usepackage{xspace}

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

One way to synchronize data stored in different formats is to use \emph{lenses},
\emph{i.e.,} programs that represent pairs of functions for translating data
from format A to B (the \emph{get} direction) and back again (the \emph{put}
direction). Lenses are usually derived using domain-specific
bidirectional languages which guarantee that the lenses defined in that language
satisfy a certain set of \emph{lens laws} that give guarantees on how the two
transformations interact.

@@

The second challenge with programming lenses in Boomerang is that even after the
programmer has specified the data as well as a canonizer that maps each member
in the source and target sets to their respective canonical representative, she
must still define the underlying lens that maps between the respective sets of
canonical representatives. This can be a slow and tedious task due to the
constraints that Boomerang imposes on its terms. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

@@

Unfortunately, programming bijective lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

@@

The first way in which our work improves upon Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe the QRE combinators and use give an example of how a
programmer might define a QRE specification of \bibtex.

@@

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}

@@

\sam{Here is my response to this. You observe that the idea of using half a lens
as a canonizer comes out of nowhere, and that this ideas is a hack and always
was. I don't think that's quite right. I think that it is useful to have a
combinator which chooses between two or more formats. Indeed, the permutation
combinator is just a special case of this. The bibtex name field is a good
example of where such a combinator would be useful.  But if we choose to have
such a combinator, then we need a way of mapping one format to another, and
really the only interesting string-to-string transformations in Boomerang are
given by lenses, which is why we use the get function of a lens in a canonizer.
We could have used any other function between the two formats but Boomerang
doesn't give us too many options for writing unidirectional transformations. So
I don't think that using a lens as a canonizer is a trick. I think that
Boomerang just isn't designed for unidirectional tranformations. But I also
agree that it doesn't make much sense to synthesize a lens only to get rid of
one half of it.
I'm not sure what the solution is to be honest.}

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp$^+$) $\to$ ",\n")
\end{lstlisting}

The final QRE combinator is the normalize combinator.
This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project (wsp+ . "," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
At this point we already have a tool for synthesizing bijective lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\sam{We don't need a general choose function. With QREs the kernel language is
always contained in the whole language so we just use the inclusion as the
choose function.}\bcp{Done.  But I wonder if the picture should mention QREs
explicitly, somehow... any ideas?}}\sam{Maybe we can put all structures, the
canonizer and the canonical strcutre in a box that says QRE?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q$ has a type $q: c_1 \Leftrightarrow c_2$ where $c_1$ and $c_2$
are QREs. In the get direction, a QRE lens $q: c_1 \Leftrightarrow c_2$ uses the
source QRE $c_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $c_1$ and then applies the $\get$ function of a
bijective lens $\ell$ to this representative. In the backward direction, $q$
operates similarly, but using the view QRE $c_2$ and the $\lput$ function of
$\ell$.

Because the QREs $c_1$ and $c_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c_1$ and $c_1$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

c & W(c) & K(c) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{c}{c_1}{f} & W(c) \sep W(c_1) & K(c_1) \\
\normalize{R}{R'}{f} & R & R' \\
c_1 \; ; \; c_2 & W(c_1) & K(c_2) \\
c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
c^* & W(c)^* & K(c)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $c = \squash{c}{c_1}{f}$, then the whole language of
$c$ is $W(c) \sep W(c_1)$. This is because the \textit{squash} combinator takes
the whole language $W(c)$ of $c$, the whole language $W(c_1)$ of $c_1$ and a
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(W(c_1))$, maps
$\mathcal{L}(W(c))$ into $\mathcal{L}(W(c_1))$ using $f$ and then canonizes
$W(c_1)$ into $K(c_1)$ using $\canonizer(c_1)$.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' \text{ for all }w, w'\\
w \; \equiv_{\squash{c_1}{c_2}{f}} \; w' &\iff& f(w) \equiv_{c_2} w'
\text{ or } w \equiv_{c_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{c_1 \; ; \; c_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(c_2)) \text{ such that } w \; \equiv_{c_1} \; k, \; w' \;
\equiv_{c_1} \; k', \text{ and } k \; \equiv_{c_2} \; k'\\
w \; \equiv_{c_1 \cdot c_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{c_1}
\; {r'}_1, \; r_2 \; \equiv{c_n} \; {r'}_2\\
w \; \equiv_{c_1 \sep c_2} \; w' &\iff& w \; \equiv_{c_1} \; w'
\text{ or } \; w \; \equiv_{c_2} \; w'\\
w \; \equiv_{c^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{c} \; {r'}_i
\\
w \; \equiv_{\perm{c_1, \ldots, c_n}{c}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{ for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{c} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Ambiguity and Well Formed QREs}
Before giving the inference rules for deriving well formed QREs, we address
the issue of {\em ambiguity}, which appears in the premises of the inference
rules for the regular combinators. If $c$ and $c_1$ are QREs, then when applying
the regular combinators to QREs, we require that if a string $s$ matches any of
the regular expressions $W(c) \cdot W(c_1)$, \quad $W(c) \sep W(c_1)$,
\quad $W(c)^*$, \quad $K(c) \cdot K(c_1)$, \quad $K(c) \sep K(c_1)$,
\quad $K(c)^*$, then $s$ matches that regular expression in only one way.

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\UnaryInfC{$\wf{\mathit{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}

\begin{prooftree}
\AxiomC{$\wf{c_i, c}$}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\wf{\perm{c_1, \ldots, c_n}{c}}$}
\end{prooftree}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$f : \mathcal{L}(W(c_1)) \longrightarrow \mathcal{L}(W(c_2))$}
\TrinaryInfC{$\wf{\squash{c_1}{c_2}{f}}$}
\DisplayProof
\end{center}
\begin{center}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

\end{center}
\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$K(c_1) = W(c_2)$}
\BinaryInfC{$\wf{c_2 \; ; \; c_1}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\wf{c}$}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$\wf{c^*}$}
\DisplayProof
\end{center}
\begin{center}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cdot^! W(c_2)$}
\AxiomC{$K(c_1) \cdot^! K(c_2)$}
\TrinaryInfC{$\wf{c_1 \cdot c_2}$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cap W(c_2) = \varnothing$}
\BinaryInfC{$\wf{c_1 \cdot c_2}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\wf{c}$}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\TrinaryInfC{$\wf{c^*}$}
\DisplayProof
\end{center}
\end{center}
\caption{Well-formed QREs
}
\label{fig:qrerules}
\end{figure}
As we already mentioned, the unambiguity conditions come into play when defining
QREs using the regular combinators. For example, the inferrence rule for
concatenation is as follows:\bcp{Rather than repeating rules inline, give
  them names in the figure.}
\begin{prooftree}
\AxiomC{$\wf{c, c_1}$}
\AxiomC{$W(c) \cdot^! W(c_1)$}
\AxiomC{$K(c) \cdot^! K(c_1)$}
\TrinaryInfC{$\wf{c \cdot c_1}$}
\end{prooftree}

>>>>

\newif\ifdraft\drafttrue  % set true to show comments
\newif\iflastminute\lastminutefalse  % for things to check at the end

@@

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption,
cleveref}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\usepackage{xspace}

@@

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonizer}{\ensuremath{\kw{canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}

\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

One way to synchronize data stored in different formats is to use \emph{lenses},
\emph{i.e.,} programs that represent pairs of functions for translating data
from format the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses are usually
derived using domain-specific bidirectional languages which guarantee that the
lenses defined in that language satisfy a certain set of \emph{lens laws} that
give guarantees on how the two transformations interact.

@@

The second challenge with programming lenses in Boomerang is that even after the
programmer has specified the data as well as a canonizer that maps each member
in the source and target sets to their respective canonical representative, she
must still define the underlying lens that maps between the respective sets of
canonical representatives. This can be a slow and tedious task due to the
constraints that Boomerang imposes on its terms. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

@@

Unfortunately, programming bijective lenses in Boomerang is often slow and
tediuos because of the fiddly constraints that Boomerang imposes on its
terms. making lens programming slow and tedious. For example, Boomerang
programmers often must rearrange the order of data items by recursively using
operators that \swap adjacent fields. Furthermore, the Boomerang type checker is
very strict, disallowing many programs because they contain ambiguity about how
certain data is transformed.

@@

The first way in which our work simplifies Boomerang's quotient lenses is
that we merge the first three pieces of quotient lenses that we just described
into a single concept, {\em Quotient Regular Expression}, or QREs. A single QRE
enables a programmer to add annotations to the outer regular expression; these
annotations can then be used to infer an inner regular expression, as well as a
canonizer from the outer regular expression to the inner one. In the following
subsection, we describe the QRE combinators and use give an example of how a
programmer might define a QRE specification of \bibtex.

@@

\begin{lstlisting}
let wsp = project [ \n\t\r]$^+$ $\to$ " "
\end{lstlisting}
\iflastminute
\bcp{Can we display blanks as a distinct ``visible blank'' symbol
(throughout)?  That will make examples easier to read.}\bcp{I tried turning
on the showstringspaces option at the top, but that doesn't do everything we
want--it only works inside quotes.  But the latex ``textvisiblespace'' command
produces a single one, and the lstlisting ``escapechar'' option will allow
us to use it inside listings.  But we should save this for the final
tweaking stage.}
% Here is an example: if you set escapechar=\%, you could use it inside the
% listings as follows: \begin{lstlisting} t %\leftarrow% 0 \end{lstlisting} 
\fi

@@

\sam{OK I suppose it's a trick though I don't see an obvious solution. }

@@

The permutation QRE combinator, \kw{perm}, allows data to be unordered. For example,
the following instance of \kw{perm} allows label, author and title fields
(which we assume have been defined earlier) to appear in any order.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title)
\end{lstlisting}
%
To normalize the field separators, one can specify that the components of the
permutation are conjoined by another QRE. For instance, below, we normalize
whitespace between fields, leaving only a single newline.
%
\begin{lstlisting}
let bib_fields = perm (label, bib_author, bib_title) with (project ("," .wsp+) $\to$ ",\n")
\end{lstlisting}

The final QRE combinator is the normalize combinator, which is also implemented
in Boomerang. This combinator allows a programmer to manually define a function
$f$ which sends each string that matches a regular expression $R$ to some
canonical representative in another regular expression $R'$. The equivalence
relation defined by the normalize combinator is hence the equivalence relation
defined by the {\em fibres} of $f$; that is, for all strings $s$ and $s'$ that
match $R$, $s$ is equivalent to $s'$ if and only if $f(s) = f(s')$.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
\label{fig:example-qre}
\end{figure}
\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\bcp{Maybe we should go a step further and label things like $K(c)$,
 $W(c)$, Canonizer(c), etc.  And probably we should display K as a subset of
 W.  I can make these changes later.)}}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: c_1 \Leftrightarrow c_2$ where $c_1$ and $c_2$
are QREs. In the get direction, a QRE lens $q_1: c_1 \Leftrightarrow c_2$ uses the
source QRE $c_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $c_1$ and then applies the $\get$ function of a
bijective lens $\ell_1$ to this representative. In the backward direction, $q_1$
operates similarly, but using the view QRE $c_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $c_1$ and $c_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective lenses is
directed by these formats, $c_1$ and $c_2$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

@@

c & W(c) & K(c) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{c}{c_1}{f} & W(c) \sep W(c_1) & K(c_1) \\
\normalize{R}{R'}{f} & R & R' \\
c_1 \; ; \; c_2 & W(c_1) & K(c_2) \\
c_1 \cdot c_2 & W(c_1) \cdot W(c_2) & K(c_1) \cdot K(c_2) \\
c_1 \sep c_2 & W(c_1) \sep W(c_2) & K(c_1) \sep K(c_2) \\
c^* & W(c)^* & K(c)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{c_1, \ldots, c_n}{c} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(c) \cdot \ldots \cdot
W(c) \cdot W(c_{\sigma(n)})\\
K( \perm{c_1, \ldots, c_n}{c} ) = & K(c_1) \cdot K(c) \cdot \ldots \cdot K(c)
\cdot K(c_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $c = \squash{c_1}{c_2}{f}$, then the whole language
of $c$ is $W(c_1) \sep W(c_2)$. This is because the \textit{squash} combinator
takes the whole language $W(c_1)$ of $c_1$, the whole language $W(c_2)$ of $c_2$
and a function $f : \mathcal{L}(W(c_1)) \longrightarrow \mathcal{L}(W(c_2))$,
maps $\mathcal{L}(W(c_1))$ into $\mathcal{L}(W(c_2))$ using $f$ and then
canonizes $W(c_2)$ into $K(c_2)$ using $\canonizer(c_2)$.

@@

\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}
w \; \equiv_R \; w' &\iff& w = w' \\
w \; \equiv_{\project{R}{s}} \; w' \text{ for all }w, w'\\
w \; \equiv_{\squash{c_1}{c_2}{f}} \; w' &\iff& f(w) \equiv_{c_2} w'
\text{ or } w \equiv_{c_2} w' \\
w \; \equiv_{\normalize{R}{R'}{f}} \; w' &\iff&
f(w)=f(w') \text{ or }w = w'\\
w \; \equiv_{c_1 \; ; \; c_2} \; w' &\iff& \exists k, k' \in
\mathcal{L}(K(c_2)) \text{ such that } w \; \equiv_{c_1} \; k, \; w' \;
\equiv_{c_1} \; k', \text{ and } k \; \equiv_{c_2} \; k'\\
w \; \equiv_{c_1 \cdot c_2} \; w'  &\iff& w = r_1
\cdot r_2, \; w' = {r'}_1 \cdot {r'}_2 \text{ with } r_1 \; \equiv_{c_1}
\; {r'}_1, \; r_2 \; \equiv{c_n} \; {r'}_2\\
w \; \equiv_{c_1 \sep c_2} \; w' &\iff& w \; \equiv_{c_1} \; w'
\text{ or } \; w \; \equiv_{c_2} \; w'\\
w \; \equiv_{c^*} \; w' &\iff& w = r_1 \cdot \ldots \cdot r_n, \; w'
= {r'}_1 \cdot \ldots \cdot {r'}_n \text{ and } r_i \equiv_{c} \; {r'}_i
\\
w \; \equiv_{\perm{c_1, \ldots, c_n}{c}} \; w' &\iff& w = r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}, \;
w' = {r'}_{\theta(1)} \cdot s'_1 \cdot \ldots \cdot s'_{n-1}
\cdot {r'}_{\theta(n)} \\
& & \text{ for some } \sigma, \theta \in S_n, \text{ with } r_i \;
\equiv_{q_i} \; r'_i \text{ and } s_k \; \equiv_{c} \; s'_{k}
\end{array}
\]
\caption{QRE Equivalence Relations}
\label{fig:relations}
\end{figure}
\subsection{Well Formed QREs}
\subsubsection{Completeness of {\em normalize}}
Before we present the full inference rules, for QREs, we discuss the {\em
normalize} QRE which is special in two main ways, the first of which has to do
with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $c$ is semantically
equivalent to $\normalize{W(c)}{K(c)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(c)) \longrightarrow \mathcal{L}(K(c))$; this is the
second property that distinguishes the {\em normalize} combinator from the other
QREs. The {\em normalize} combinator hence gives a sufficient condition for
potential QREs to be ``well-behaved'', particularly with respect to the
strategy for synthesizing quotient lenses which we shall introduce in
Section~\ref{synth}. This is the
\subsubsection{Ambiguity}
Another issue which we address before presenting the full set of inference
rules for QREs, we discuss the issue of {\em ambiguity}, which appears in the
premises of the inference rules for the regular combinators. If $c_1$ and $c_2$
are QREs, then when applying the regular combinators to QREs, we require that
if a string $s$ matches any of the regular expressions $W(c_1) \cdot W(c_2)$,
\quad $W(c_1) \sep W(c_2)$, \quad $W(c_1)^*$, \quad $K(c_1) \cdot K(c_2)$,
\quad $K(c_1) \sep K(c_2)$, \quad $K(c_1)^*$, then $s$ matches that regular
expression in only one way.

@@

Figure~\ref{fig:qrerules} gives the inferrence rules for deriving well-formed
QREs.
\begin{figure}[tp!]
\begin{center}
\AxiomC{$R$ is strongly unambiguous}
\RightLabel{(Id)}
\UnaryInfC{$\wf{\mathit{id}(R)}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$s \in \mathcal{L}(R)$}
\RightLabel{(Project)}
\UnaryInfC{$\wf{\project{R}{s}}$}
\DisplayProof
\hskip 1.5em
\end{center}

\begin{prooftree}
\AxiomC{$\wf{c_i, c}$}
\RightLabel{(Perm)}
\AxiomC{${\substack{\forall \sigma \neq \theta, \; W(c_{\sigma(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\sigma(n)})\\ \cap W(c_{\theta(1)}) \cdot W(c)
\cdot \ldots \cdot W(c_{\theta(n)}) =\varnothing}}$}
\AxiomC{$K(c_1) \cdot^! K(c) \cdot^! \ldots \cdot^! K(c) \cdot^! K(c_n)$}
\TrinaryInfC{$\wf{\perm{c_1, \ldots, c_n}{c}}$}
\end{prooftree}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$f : \mathcal{L}(W(c_1)) \longrightarrow \mathcal{L}(W(c_2))$}
\RightLabel{(Squash)}
\TrinaryInfC{$\wf{\squash{c_1}{c_2}{f}}$}
\DisplayProof
\end{center}
\begin{center}

\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

\end{center}
\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$K(c_1) = W(c_2)$}
\RightLabel{(Compose)}
\BinaryInfC{$\wf{c_1 \; ; \; c_2}$}
\DisplayProof
\hskip 1.5em
\AxiomC{$\wf{c}$}
\AxiomC{$W(c)^{*!}$}
\AxiomC{$K(c)^{*!}$}
\RightLabel{(Star)}
\TrinaryInfC{$\wf{c^*}$}
\DisplayProof
\end{center}
\begin{center}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cdot^! W(c_2)$}
\AxiomC{$K(c_1) \cdot^! K(c_2)$}
\RightLabel{(Concat)}
\TrinaryInfC{$\wf{c_1 \cdot c_2}$}
\DisplayProof
\end{center}

\begin{center}
\AxiomC{$\wf{c_1, c_2}$}
\AxiomC{$W(c_1) \cap W(c_2) = \varnothing$}
\RightLabel{(Union)}
\BinaryInfC{$\wf{c_1 \sep c_2}$}
\DisplayProof
\
\end{center}
\end{center}
\caption{Well-formed QREs}
\label{fig:qrerules}
\end{figure}
As we already mentioned, the unambiguity conditions come into play when defining
QREs using the regular combinators. For example, the (Concat) inferrence rule
for concatenation says that the concatenation $c_1 \cdot c_2$ of QREs $c_1$ and
$c_2$ is well formed only if the concatentions of $W(c_1)$ and $W(c_2)$, and
$K(c_1)$ and $K(c_2)$ are unambiguous.
